---
title: "Species Distribution Modeling"
output: html_document
---

I am modeling the distributions of the four *Enyalius* study species (*E. catenatus*, *E. pictus*, *E. perditus*, and *E. iheringii*) to understand their current distributions and responses to long-term historical climate change.  

I will build robust models of their current distributions using recent environmental data (i.e. last few decades). Then, I will build similar models with variables that are available for both present-day and historical (i.e. last glacial maximum) time periods. I will use the **current** models as benchmarks for the performance of the models intended for hindcasting.  

```{r setup}
library(sf)
library(terra)
library(janitor)
library(here)
library(rnaturalearth)
library(rasterVis)
library(tidyverse)
```


# Current

## Variable selection

I'm choosing variables that encompass the range of the *Enyalius* species, are relevant to their physiology, and are obtainable at a reasonable spatial resolution. 

Bioclims:  
Using the BrazilClim dataset downloaded from Ramoni-Perazzi, P., Passamani, M., Thielen, D., Padovani, C., Arizapana-Almonacid, M.A., 2021. BrazilClim : The overcoming of limitations of pre‐existing bioclimate data. Int. J. Climatol. https://doi.org/10.1002/joc.7325. Downloaded on __2022-02-25__.  
1 km resolution.  

Land Cover:  
Using the forest land cover rasters (classes 1-4) from Tuanmu, M.-N. and W. Jetz. 2014. A global 1-km consensus land-cover product for biodiversity and ecosystem modeling. Global Ecology and Biogeography 23(9): 1031-1045. Downloaded on __2022-02-26__.  
1 km resolution  

## Variable processing

I'm cropping the variables to a relevant extent before going further with the individual SDMs. I'm cropping them with a 1 degree buffer around the entire species group's range. 

First, I'm reading in the localities and plotting them to make sure there isn't anything wild. Everything looks reasonable!  

```{r, locs, message=FALSE}
# read in as a data frame first
locs_df <- read_csv(here("analysis", "data", "enyalius_locs.csv")) %>% 
  # the variable names are messy
  janitor::clean_names() %>%
  filter(!is.na(latitude), !is.na(longitude),
         # remove localities that are only mapped to a Google Earth municipality. The reserve in the "GPS of the reserve" locality is barely a km, so that is acceptable error relative to the environmental resolution
         source_lat_long != "Google Earth municipality") %>%
  # there are some lat/longs with spaces. Need to fix this
  mutate(longitude = str_remove(longitude, "\\s"), 
         latitude = str_remove(latitude, "\\s")) %>% 
  # convert to numeric
  mutate(longitude = as.numeric(longitude),
         latitude = as.numeric(latitude))

# convert to sf for spatial operations
locs_sf <- st_as_sf(locs_df, 
                    coords = c("longitude", "latitude"),
                    crs = 4326)

# convert to vect for interacting with terra
locs_vec <- vect(locs_sf)

# atlantic forest shapefile for plotting
af <- read_sf(here("analysis", "data", "atlantic_forest", "atlantic_forest.geojson"))

# plot
ggplot() +
  geom_sf(data = st_geometry(af)) +
  geom_sf(data = locs_sf)
```

Creating a convex hull with a 1 degree buffer.  

```{r mcp-all, warning=FALSE}
mcp_all <- st_convex_hull(st_union(locs_sf)) %>%
  st_buffer(dist = 1.0) %>% 
  terra::vect()

plot(st_geometry(af))
plot(mcp_all, add=TRUE)
plot(locs_vec, add=TRUE)
```

Crop bioclims.  

```{r crop-env, eval=FALSE}
bioclims <- terra::rast(list.files(here("analysis", "data", "brazil_clim"), full.names = TRUE)) %>% 
  terra::crop(mcp_all) %>% 
  terra::mask(mcp_all)

# fix the names  
names(bioclims) <- str_remove(names(bioclims), "BC1.0_")


plot(bioclims$bio_01)
plot(st_geometry(af), add=TRUE)
plot(mcp_all, add=TRUE)
plot(locs_vec, add=TRUE)
```

Crop land cover.  

```{r}
landcover <- terra::rast(list.files(here("analysis", "data", "brazil_clim"), full.names = TRUE)) %>% 
  terra::crop(mcp_all) %>% 
  terra::mask(mcp_all)

# fix the names  
names(bioclims) <- str_remove(names(bioclims), "BC1.0_")


plot(bioclims$bio_01)
plot(st_geometry(af), add=TRUE)
plot(mcp_all, add=TRUE)
plot(locs_vec, add=TRUE)
```



## E. catenatus


## E. pictus


## E. perditus


## E. iheringii
