---
title: "Assembly QC"
format: 
  html:
    code-fold: true
editor: visual
---

This report outlines quality control (QC) assessments performed on *Enyalius* `ipyrad` assemblies, with a focus on determining the appropriate clustering threshold. Thresholds of 0.90, 0.92, 0.94, 0.96, and 0.98 were considered. To determine the impact of batch effects, we are:

-   plotting heterozygosity per plate, per species

-   examining missingness per plate, per species

-   including a visualization of a PCA that includes the plate number per species

For the batch effect analysis, we're arbitrarily using the cluster threshold of 0.94. These values may vary a little bit from the optimal clustering threshold to be determined, but shouldn't impact any conclusions.

To determine the appropriate clustering threshold, we are following [McCartney-Melstad et al. 2019](https://onlinelibrary.wiley.com/doi/10.1111/1755-0998.13029)'s workflow, which includes the following steps:

-   Genetic diversity (heterozygosity \~ number of SNPs recovered) 

-   Cumulative variance explained by major principal components

-   Relationship between missingness and genetic divergence

-   Slope of isolation by distance

-   Phylogenetic resolution

Also, as a sanity check, the error rate should be on the order of 0.001 and heterozygosity should be on the order of 0.01.

Note, *ipyrad* no longer outputs paralog flags, so we are not performing the original first step, visualizing the fraction of inferred paralogs.

```{r}
#| label: r-setup
#| warning: false
#| message: false

library(here)
library(tidyverse)
library(reticulate)

source(here("assembly", "scripts", "read_ipyrad_stats.R"))

use_condaenv(condaenv = "enyaliuspy")

# read in localities
locs <- read_csv(here("analysis", "data", "enyalius_locs.csv"))

```

```{python}
#| label: python-setup
#| warning: false
#| message: false

import ipyrad
import ipyrad.analysis as ipa
from ipyrad.analysis.sharing import Sharing
import pandas as pd
import toyplot
import matplotlib.pyplot as plt


```

::: panel-tabset
## Batch effects

The *Enyalius* individuals were sequenced across multiple plates, which can result in non-biological variation among samples due to library prep and sequencing artifacts. Here, I'm visualizing heterozygosity, missingness, and genetic structure (PCA) patterns across sequencing plates and species.

```{r}
#| label: batch-read-data
#| message: false
#| warning: false

ip_stats_94 <- read_ipyrad_stats(here("assembly", "full", "clust94_outfiles", "clust94_stats.txt"))

locs_stats_94 <- left_join(
  ip_stats_94, 
  locs,
  by = "id_code"
) %>% 
  # remove and rename duplicate columns
  select(-ends_with(".y")) %>% 
  rename_with(~str_remove(.x, ".x"), ends_with(".x")) %>% 
  # there are two concatenated samples that don't have plate info
  # don't want outgroups
  filter(!is.na(plate),
         species %in% c("cat", "ihe", "per", "pic")) 

# snps filepath for missingness matrix
snps_path_94 <- here("assembly", "full", "clust94_outfiles", "clust94.snps.hdf5")

```

::: panel-tabset
## Heterozygosity

Heterozygosity varies across plates, but species mostly correspond with a single plate. For species sequenced across multiple plates, there doesn't appear to be plate effects.

Error varies across plates in general and within species. We'll need to consider this for future analyses and may be able to account for it bioinformatically

```{r}
#| label: heterozygosity-plots
#| layout-ncol: 2

locs_stats_94 %>% 
  ggplot(aes(x = hetero_est, y = plate)) + 
  geom_boxplot() +
  geom_jitter(height = 0.2) +
  scale_color_viridis_d() +
  labs(
    y = "", 
    x = "Heterozygosity est.",
    title = "Heterozygosity") +
  theme_bw()

locs_stats_94 %>% 
  ggplot(aes(x = hetero_est, fill = plate)) + 
  geom_dotplot(binwidth = 0.0002) +
  scale_fill_viridis_d() +
  labs(y = "", x = "Heterozygosity est.") +
  facet_wrap(~species) +
  theme_bw() +
  theme(legend.position = "top")
```

```{r}
#| label: error-plots
#| layout-ncol: 2

locs_stats_94 %>% 
  ggplot(aes(x = error_est, y = plate)) + 
  geom_boxplot() +
  geom_jitter(height = 0.2) +
  scale_color_viridis_d() +
  labs(
    y = "", 
    x = "Error est.",
    title = "Error") +
  theme_bw()

locs_stats_94 %>% 
  ggplot(aes(x = error_est, fill = plate)) + 
  geom_dotplot(binwidth = 3e-5) +
  scale_fill_viridis_d() +
  labs(y = "", x = "Error est.") +
  facet_wrap(~species) +
  theme_bw() +
  theme(legend.position = "top")

```

## Missingness

I'm using *ipyrad* to calculate allele sharing among samples and visualize as a heat map. Alleles should be shared most within species and shouldn't be structured by plate number.

In the first iteration, there are a couple outlier individuals (*per_bana_CTMZ03943* and *cat_cam_PEU322*). These individuals also had a low number of reads and recovered loci. We have multiple individuals from each locality, so we're getting rid of them for future analyses.

```{python}
#| label: sharing-calculation
#| message: false
#| output: false
#| cache: true

# this takes a while. I'd write it to file and read it in if you plan on running this multiple times
share = Sharing(
    data=r.snps_path_94
)
share.run()

```

```{r}
#| label: sharing-plot
heatmap(
  as.matrix(py$share$sharing_matrix),
  Rowv = NA,
  Colv = NA,
  symm = TRUE,
  main = "Proportion of shared alleles",
  col= colorRampPalette(RColorBrewer::brewer.pal(8, "Oranges"))(25)
  )

legend(x="left", legend=c("low", "medium", "high"), 
     fill=colorRampPalette(RColorBrewer::brewer.pal(8, "Oranges"))(3))
```

```{r}
#| label: sharing-plot-no-outliers

sdf <- py$share$sharing_matrix %>% 
  as_tibble(rownames = "sample") %>% 
  select(-per_bana_CTMZ03943, -cat_cam_PEU322) %>% 
  filter(sample != "per_bana_CTMZ03943", sample != "cat_cam_PEU322") 

sm <- as.matrix(sdf %>% select(-sample))
rownames(sm) <- sdf$sample

heatmap(
  sm,
  Rowv = NA,
  Colv = NA,
  symm = TRUE,
  main = "Proportion of shared alleles, no outliers",
  col= colorRampPalette(RColorBrewer::brewer.pal(8, "Oranges"))(25)
  )

legend(x="left", legend=c("low", "medium", "high"), 
     fill=colorRampPalette(RColorBrewer::brewer.pal(8, "Oranges"))(3))
```

There appears to be some within-species variation, but it's not due to a batch effect. We visualized *E. pictus* and *E. iheringii* because they have individuals spread across plates.

```{r}
#| label: pictus-iheringii-outliers
#| layout-ncol: 1

locs_sdf_94 <- left_join(
  sdf %>% rename(id_code = sample), 
  locs,
  by = "id_code"
)

locs_sdf_94 %>% 
  filter(species == "pictus") %>% 
  select(starts_with("pic"), plate) %>% 
  pivot_longer(starts_with("pic"), names_to = "sample", values_to = "num_shared") %>% 
  ggplot(aes(x = num_shared, fill = plate)) + 
  geom_dotplot(binwidth = 6500) +
  labs(x = "Number of shared alleles",
       y = "",
       title = "Number of shared alleles for E. pictus",
       caption = "The zero values are due to the diagonal in the sharing matrix.") +
  theme_bw() +
  theme(legend.position = "bottom")
  
locs_sdf_94 %>% 
  filter(species == "iheringii") %>% 
  select(starts_with("ihe"), plate) %>% 
  pivot_longer(starts_with("ihe"), names_to = "sample", values_to = "num_shared") %>% 
  ggplot(aes(x = num_shared, fill = plate)) + 
  geom_dotplot(binwidth = 3000) +
  labs(x = "Number of shared alleles",
       y = "",
       title = "Number of shared alleles for E. iheringii",
       caption = "The zero values are due to the diagonal in the sharing matrix.") +
  theme_bw() +
  theme(legend.position = "bottom")

```

## PCA
:::

## Cluster evaluation
:::

