---
title: "ABC Analysis"
format: html
editor: visual
---

# Approximate Bayesian Computation Analysis

## Setup

```{r}
#| label: load-packages
#| warning: false
#| message: false

library(tidyverse)
library(abc)
library(here)

```

```{r}
#| label: read-data
#| warning: false
#| message: false

# read in sumstats
## iheringii
### Simulations
lin_ihe_paths <- list.files(here("analysis", "output", "simulations"), pattern = "linear", full.names = TRUE)
lin_ihe <- map_df(lin_ihe_paths, read_csv) %>% 
  mutate(transformation = "linear")

hinge_ihe <- read_csv(here("analysis", "output", "simulations", "sims_2023-11-20_hinge.csv")) %>% 
  mutate(transformation = "hinge")

sims_ihe <- bind_rows(lin_ihe, hinge_ihe) %>% 
  # fill in the missing param values
  group_by(param_id) %>% 
  fill(ancestral_n, total_inds_curr) %>% 
  fill(ancestral_n, total_inds_curr, .direction = "up") %>% 
  ungroup()
  

### Empirical
empirical_ihe <- read_csv(here("analysis", "output", "empirical_sumstats", "iheringii_sumstats.csv")) %>% 
  dplyr::select(-num_var)

## catenatus

```

## Exploration

### PCA

#### *E. iheringii*

```{r}
#| label: pca-ihe

pca_df_ihe <- sims_ihe %>% 
  dplyr::select(
    starts_with("sfs"),
    "taj_d",
    starts_with("pi"),
    starts_with("ibd"),
    "morans_i"
  )

pca_ihe <- prcomp(pca_df_ihe, center=TRUE, scale. = FALSE)

pca_df_ihe <- bind_cols(pca_ihe$x, sims_ihe)

ggplot(pca_df_ihe, aes(x = PC1, y = PC2, color = transformation)) +
  geom_point(alpha = 0.5) +
  scale_color_viridis_c() +
  theme_bw()
```

## ABC

### Model Selection

#### *E. iheringii*

Model selection to assess discriminatory ability among simulations.

```{r}
#| label: model-selection-ihe
#| cache: true

sumstats_ihe <- sims_ihe %>% 
  dplyr::select(
    starts_with("sfs"),
    "taj_d",
    starts_with("pi"),
    starts_with("ibd"),
    "morans_i"
  )

cv_modsel_ihe <- cv4postpr(sims_ihe$transformation, sumstats_ihe, nval = 500, tol = 0.05, method="mnlogistic")

summary(cv_modsel_ihe)
```

Model selection using the empirical data and a rejection algorithm.

```{r}
#| label: model-selection-rejection-ihe

modsel_emp_ihe <- postpr(empirical_ihe, sims_ihe$transformation, sumstats_ihe, tol=.05, method="mnlogistic")

summary(modsel_emp_ihe)

```

### Goodness-of-fit

#### *E. iheringii*

PCA

```{r}
#| label: gof-pca-ihe

gfitpca(empirical_ihe, sumstat=sumstats_ihe, index=sims_ihe$transformation, cprob=.1)

```

Goodness-of-fit test linear.

```{r}
#| label: gof-test-lin-ihe
#| cache: true

ss_lin_ihe <- bind_cols(sumstats_ihe, transformation = sims_ihe$transformation) %>% 
  filter(transformation == "linear") %>% 
  dplyr::select(-transformation)

gof_linear <- gfit(target=empirical_ihe, sumstat=ss_lin_ihe, statistic=mean, tol = 0.05, nb.replicate=100)

plot(gof_linear, main="Histogram under linear model")

```

Goodness-of-fit test hinge.

```{r}
#| label: gof-test-hinge-ihe
#| cache: true

ss_hinge_ihe <- bind_cols(sumstats_ihe, transformation = sims_ihe$transformation) %>% 
  filter(transformation == "hinge") %>% 
  dplyr::select(-transformation)

gof_hinge <- gfit(target=empirical_ihe, sumstat=ss_hinge_ihe, statistic=mean, tol = 0.05, nb.replicate=100)

plot(gof_hinge, main="Histogram under threshold model")

```

Empirical summary statistics

```{r}
empirical_ihe %>% knitr::kable(digits = 3)
```
