---
title: "Species Distribution Modeling"
output:
  html_document:
    toc: true
    toc_float: 
      toc_depth: 4
      collapsed: false
editor: visual
knitr:
  opts_chunk:
    warning: false
    message: false
---

I am modeling the distributions of the four *Enyalius* study species (*E. catenatus*, *E. pictus*, *E. perditus*, and *E. iheringii*) to understand their current distributions and responses to long-term historical climate change.

I will build robust models of their current distributions using recent environmental data (i.e. last few decades). Then, I will build similar models with variables that are available for both present-day and historical (i.e. last glacial maximum) time periods. I will use the **current** models as benchmarks for the performance of the models intended for hindcasting.

```{r}
#| label: setup
#| message: false
#| warning: false

library(sf)
library(spThin)
library(janitor)
library(here)
library(rasterVis)
library(tidyverse)
library(ENMeval)
library(patchwork)
library(terra)
library(usdm)
```

# Current

## Variable selection

I'm choosing variables that encompass the range of the *Enyalius* species, are relevant to their physiology, and are obtainable at a reasonable spatial resolution.

Bioclims:  
Using the BrazilClim dataset downloaded from Ramoni-Perazzi, P., Passamani, M., Thielen, D., Padovani, C., Arizapana-Almonacid, M.A., 2021. BrazilClim : The overcoming of limitations of pre‐existing bioclimate data. Int. J. Climatol. <https://doi.org/10.1002/joc.7325>. Downloaded on **2022-02-25**.  
1 km resolution.

For projecting to past climates, I'm using [CHELSA TraCE21k v1.0 Bioclims](https://cp.copernicus.org/preprints/cp-2021-30/cp-2021-30.pdf) at 100-year intervals back to the LGM. 1 km resolution. **Not downloaded yet- need to fill this in when it happens**

Land Cover:  
Using biome projections from [Costa et al. 2018](https://onlinelibrary-wiley-com.ezproxy.gc.cuny.edu/doi/full/10.1111/geb.12694) at 1000-year intervals back to the LGM. 5 km resolution that I'm downscaling with nearest-neighbor resampling. Downloaded on **2023-06-08**.

## Variable processing

I'm cropping the variables to a relevant extent before going further with the individual SDMs. I'm cropping them with a 1 degree buffer around the entire species group's range.

First, I'm reading in the localities and plotting them to make sure there isn't anything wild. Everything looks reasonable!

```{r}
#| label: locs
#| message: false
#| eval: false
  
  
# read in as a data frame first
locs_df <- read_csv(here("analysis", "data", "enyalius_locs.csv")) %>% 
  # the variable names are messy
  janitor::clean_names() %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  # there are some lat/longs with spaces. Need to fix this
  mutate(longitude = str_remove(longitude, "\\s"), 
         latitude = str_remove(latitude, "\\s")) %>% 
  # convert to numeric
  mutate(longitude = as.numeric(longitude),
         latitude = as.numeric(latitude))

# convert to sf for spatial operations
locs_sf <- st_as_sf(locs_df, 
                    coords = c("longitude", "latitude"),
                    crs = 4326)

# convert to vect for interacting with terra
locs_vec <- vect(locs_sf)

# atlantic forest shapefile for plotting
af <- read_sf(here("analysis", "data", "atlantic_forest", "atlantic_forest.geojson"))

# plot
ggplot() +
  geom_sf(data = st_geometry(af), fill = "gray") +
  geom_sf(data = locs_sf, aes(fill = species), shape = 21) +
  scale_fill_viridis_d() +
  theme_minimal()
```

Creating a convex hull with an 2 degree buffer.

```{r}
#| label: mcp-all
#| warning: false
#| eval: false

mcp_all <- st_convex_hull(st_union(locs_sf)) %>%
  st_buffer(dist = units::set_units(2, degree)) %>% 
  terra::vect()

plot(st_geometry(af))
plot(mcp_all, add=TRUE)
plot(locs_vec, add=TRUE)
```

Read in bioclims. Originally I waited to crop, but I folded that into the initial variable downloads. The rasters going into `cropped_predictors` are the same as I'm reading in, just with different names. They're also being written as a single file, rather than multiple files.

```{r}
#| label: crop-env
#| warning: false
#| eval: false

bioclims <- terra::rast(list.files(here("analysis", "data", "current_climate_chelsa"), full.names = TRUE))

# get the only the bioclim labels
names(bioclims) <- str_extract(names(bioclims), "bio[\\d+]*")

plot(bioclims$bio1)
plot(st_geometry(af), add=TRUE)
plot(mcp_all, add=TRUE)
plot(locs_vec, add=TRUE)
```

Crop the present-day biome layer and convert it to a binary variable, where 1 is the presence of tropical moist forest and 0 is the absence of tropical moist forest.

```{r}
#| label: forest-cover
#| warning: false
#| eval: false

forest_cover <- terra::rast(here("analysis", "data", "biomes", "S1biomes_0k.asc")) %>% 
  aggregate(2) %>% 
  terra::crop(mcp_all) %>% 
  terra::mask(mcp_all) 

forest_cover[forest_cover != 1] <- 0

names(forest_cover) <- "forest_cover"


plot(forest_cover)
plot(st_geometry(af), add=TRUE)
plot(mcp_all, add=TRUE)
plot(locs_vec, add=TRUE)
```

Write the cropped climate and forest cover layers to file for use in the SDMs.

```{r}
#| label: write-cropped
#| warning: false
#| eval: false

terra::writeRaster(bioclims, here("analysis", "output", "cropped_predictors", "bioclims.tif"))
terra::writeRaster(forest_cover, here("analysis", "output", "cropped_predictors", "forest_cover.tif"))
```

## General SDM steps {.tabset}

For each SDM, I'm going to perform the following steps:

```         
- spatially thin localities  
- crop the new predictors according to a minimum convex hull of the localities for each species with a 0.5 degree buffer 
    - I tried point-buffers, but they resulted in worse models
- extract background environmental data (10,000 points) for predictor correlation and modeling
- correlate predictors
- remove predictors w/ r > 0.75, prioritizing ecologically relevant variables  
- Use Maxent for modeling
    - LQH feature classes to keep models relatively simple  
    - regularization multipliers from 0.5 to 5.0 in 0.5 increments to test a wide range of regularization
    - leave-one-out cross validation for model selection due to a low number of localities
    - select models first by AICc (model fit), followed by ommission error rate (prediction)  
```

In addition, I'm reading/writing data for each species in isolation, so the code for one species isn't dependent on that for another, or for what I did during variable processing.

### E. catenatus

Note: putative "catenatus 2" species from Mariana's phylogenetic work have been removed.

Reading in data for plotting.

```{r}
#| label: read-af

# atlantic forest shapefile
af <- read_sf(here("analysis", "data", "atlantic_forest", "atlantic_forest.geojson"))
```

#### Spatial thin

Read and filter localities.

```{r}
#| label: locs-cat

locs_cat <- read_csv(here("analysis", "data", "enyalius_locs.csv")) %>% 
  # the variable names are messy
  janitor::clean_names() %>%
  filter(species == "catenatus",
    !is.na(latitude), !is.na(longitude),
    # remove duplicates
    !duplicated(latitude)) %>%
  # there are some lat/longs with spaces. Need to fix this
  mutate(longitude = str_remove(longitude, "\\s"), 
         latitude = str_remove(latitude, "\\s")) %>% 
  # convert to numeric
  mutate(longitude = as.numeric(longitude),
         latitude = as.numeric(latitude)) %>% 
  # one locality falls into the ocean when using the aggregated data. Have to scooch it over by 0.05 degrees to overlap with the correct cell
  mutate(longitude = if_else(longitude == -39.06, longitude - 0.05, longitude)) %>% 
  # convert to sf for spatial operations
  st_as_sf(coords = c("longitude", "latitude"),
           crs = 4326,
           remove = FALSE)

# plot localities
plot(st_geometry(af))
plot(st_geometry(locs_cat), add = TRUE)
```

Spatial thin. I'm using a 20 km buffer to reduce the impact of spatial autocorrelation.

```{r}
#| label: spthin-cat

set.seed(394833)

#run spthin algorithm. This returns 100 possible combinations of removed localities
output <-
  spThin::thin(
    locs_cat,
    'latitude',
    'longitude',
    'species',
    thin.par = 20,
    reps = 100,
    locs.thinned.list.return = TRUE,
    write.files = FALSE,
    verbose = FALSE
  )

# I want to maximize the # of localities returned  
maxThin <- which(sapply(output, nrow) == max(sapply(output, nrow)))

# if there are multiple iterations with the max # of localities, pick one
maxThin <-
  output[[ifelse(length(maxThin) > 1, maxThin[1], maxThin)]]

# subset locs to match only thinned locs
locs_cat <- locs_cat[as.numeric(rownames(maxThin)), ]

# get the unused locs as a testing data set
# this probably isn't useful since they will overlap heavily with the training data, but seems like another piece of info to look at
test_cat <- locs_cat[-as.numeric(rownames(maxThin)), ]

plot(st_geometry(af))
plot(st_geometry(locs_cat), add=TRUE)
```

Write thinned localities to file

```{r, eval=FALSE}
#| label: write-spthin-cat
#| eval: false

# Write to file
st_write(locs_cat, here("analysis", "output", "thinned_localities", "catenatus_thinned.gpkg"))
```

#### Crop environment

First, I need to read in the environmental data.

```{r}
#| label: read-envt-cat

bioclims <- terra::rast(here("analysis", "output", "cropped_predictors", "bioclims.tif"))
forest_cover <- terra::rast(here("analysis", "output", "cropped_predictors", "forest_cover.tif"))
```

Cropping and combining variables for analysis.

```{r}
#| label: crop-cat
#| layout-ncol: 2
#| warning: FALSE

# for projection
mcp_cat <- st_convex_hull(st_union(locs_cat)) %>%
  st_buffer(dist = units::set_units(0.5, degree)) %>% 
  terra::vect()

bioclims_cat <- bioclims %>%
  # the bioclims are in a slightly different CRS
  terra::project(forest_cover) %>% 
  terra::crop(mcp_cat) %>% 
  terra::mask(mcp_cat)

forest_cover_cat <- forest_cover %>% 
  terra::crop(bioclims_cat) %>% 
  terra::mask(bioclims_cat[[1]]) 

predictors_cat <- c(bioclims_cat)

plot(bioclims_cat[[1]])
plot(forest_cover_cat)
plot(locs_cat, add=TRUE)
```

#### Predictor correlation

Sample 10000 background points. Only returned 2355 points.

```{r}
#| label: bg-cat
set.seed(1988888)

# for variable correlations
bg_envt_cat <- terra::spatSample(predictors_cat, 10000, 
                               warn=TRUE, 
                               na.rm = TRUE, 
                               as.df = TRUE,
                               xy = TRUE)

# for use in ENMevaluate
bg_coords_cat <- bg_envt_cat[,c("x", "y")]
```

Next, I'll extract the values for the background points and perform variance inflation factor stepwise selection with a VIF threshold of 10.

```{r}
#| label: vif-cat
# extract values
bg_corr_cat <- bg_envt_cat %>% select(-x, -y)

usdm::vifstep(bg_corr_cat, th=10)
```

The final variable list: BIO3, BIO4, BIO8, BIO13, BIO15, BIO18

```{r}
# label: pred-cat

predictors_cat <- predictors_cat[[c("bio3", "bio4", "bio8", "bio13", "bio15", "bio18")]]
```

#### Maxent model

I'm using a jackknifing model evaluation approach since I only have 26 observations and a previous attempt for spatial CV led to wonky evaluation metrics. Additionally, spatial CV can lead to worse predictions than non-spatial CV: [Wadoux et al. 2021](https://www-sciencedirect-com.ezproxy.gc.cuny.edu/science/article/pii/S0304380021002489).

```{r}
#| label: folds-cat
set.seed(7990777)
coords_cat <- st_coordinates(locs_cat)
colnames(coords_cat) <- c("x", "y")

folds_cat <- ENMeval::get.jackknife(occ = coords_cat, 
                            bg = bg_coords_cat)
```

Run the model. Predictions are clamped to prevent extrapolation.

```{r}
#| label: run-model-cat
#| eval: false

set.seed(34622)

# the vector of regularization multipliers to test
rms <- seq(0.5, 5, 0.5)

# convert the terra object to a raster stack for use in EMNeval
predictors_cat_stack <- raster::stack(predictors_cat)

# iterate model building over all chosen parameter settings
sdm_cat <-
  ENMeval::ENMevaluate(
    occs = coords_cat,
    envs = predictors_cat_stack,
    bg.coords = bg_coords_cat,
    RMvalues = rms,
    fc = c('L', 'LQ'),
    # clamping to prevent model extrapolation
    doClamp = TRUE,
    taxon.name = "catenatus",
    partitions = "user",
    user.grp = folds_cat,
    bg.grp,
    #categoricals = "forest_cover",
    clamp = TRUE,
    algorithm = "maxent.jar",
    parallel = TRUE,
    numCores = 6
  )
```

```{r}
#| label: write-cat-model
#| eval: false

# write the model to file
write_rds(sdm_cat, here("analysis", "output", "sdm_models", "sdm_catenatus.rds"))
```

```{r}
#| label: read-cat-model
#| echo: false

sdm_cat <- read_rds(here("analysis", "output", "sdm_models", "sdm_catenatus.rds"))
```

##### Model evaluation

Let's take a look at the model results.

```{r}
#| label: mod-results-cat
eval_table_cat <- sdm_cat@results
eval_mods_cat <- sdm_cat@models

names(eval_mods_cat) <-
  str_replace_all(names(eval_mods_cat), "\\.", "\\_")
```

Select the final model. First I'm looking at plots of model evaluation stats to get an idea of the spread of stats.

```{r, warning=FALSE}
#| label: eval-met-cat
daic_cat <- ggplot(data = eval_table_cat, aes(x = rm, y = delta.AICc, color = fc)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw()

or_cat <- ggplot(data = eval_table_cat, aes(x = rm, y = or.10p.avg, color = fc)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw()

dauc_cat <- ggplot(data = eval_table_cat, aes(x = rm, y = auc.diff.avg, color = fc)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw()

(daic_cat + or_cat) / (dauc_cat)
```

Now I'm going to take a look at tables of delta AICc, omission rate, and AUC to see how close the models are. The model with the lowest AICc has a regularization multiplier of 1 and L feature class.

```{r}
#| label: eval-table-cat

eval_table_cat %>% 
  select(delta.AICc, AICc, or.10p.avg, or.mtp.avg, auc.diff.avg, auc.val.avg, rm, fc) %>%
  arrange(delta.AICc) %>% 
  head(10) %>% 
  knitr::kable()
```

Select model.

```{r}
#| label: select-model-cat
mod_cat <- eval_mods_cat$rm_0_5_fc_L
opt_seq_tune_cat <- "rm.0.5_fc.L"
```

Plot the variable contributions and response curves. The most important variables are bio15, bio18, and bio13. All precipitation!

```{r}
#| label: var-importance-cat
plot(mod_cat)
```

```{r, eval=FALSE}
png("variable_contribution_catenatus.png")
plot(mod_cat)
dev.off()

file.copy(here("variable_contribution_catenatus.png"), here("analysis", "output", "sdm_response_curves_variable_importance", "variable_contribution_catenatus.png"))
file.remove(here("variable_contribution_catenatus.png"))
```

Suitability increases with precipitation amount of the wettest month, decreases with precipitation seasonality, and increases with mean monthly precipitation of the warmest quarter.

```{r}
#| label: resp-curve-cat
dismo::response(mod_cat)
```

```{r, eval=FALSE}
png("response_curves_catenatus.png")
dismo::response(mod_cat)
dev.off()

file.copy(here("response_curves_catenatus.png"), here("analysis", "output", "sdm_response_curves_variable_importance", "response_curves_catenatus.png"))
file.remove(here("response_curves_catenatus.png"))
```

##### Project

I'm projecting the model to the study area extent.

```{r}
#| label: proj-cur-cat
pred_cat <- ENMeval::eval.predictions(sdm_cat)[[opt_seq_tune_cat]]
plot(pred_cat)
plot(st_geometry(af), add = TRUE)
plot(st_geometry(locs_cat), pch = 21, bg = alpha("lightgray", 0.5), add = TRUE)
```

```{r}
#| label: proj-cur-cat-thresh
pred_cat_thresh <- pred_cat$rm.0.5_fc.L
# observed suitabilities
obs_suit_cat <- terra::extract(pred_cat$rm.0.5_fc.L, locs_cat)
# minimum training presence
min_suit_cat <- min(obs_suit_cat)

pred_cat_thresh[pred_cat_thresh >= min_suit_cat] <- 1
pred_cat_thresh[pred_cat_thresh < min_suit_cat] <- 0

plot(pred_cat_thresh)
plot(st_geometry(af), add = TRUE)
plot(st_geometry(locs_cat), pch = 21, bg = alpha("lightgray", 0.5), add = TRUE)
```

### E. pictus

Reading in data for plotting.

```{r}
#| label: af-pictus
# atlantic forest shapefile
af <- read_sf(here("analysis", "data", "atlantic_forest", "atlantic_forest.geojson"))
```

#### Spatial thin

Read and filter localities. I'm including the original genetic localities from Mariana and iNaturalist observations I downloaded on 2023-07-10. Citation: GBIF.org (10 July 2023) GBIF Occurrence Download <https://doi.org/10.15468/dl.uj2ft2>.

```{r}
#| label: locs-pic

locs_pic <- read_csv(here("analysis", "data", "enyalius_locs.csv")) %>% 
  # the variable names are messy
  janitor::clean_names() %>%
  filter(species == "pictus",
    !is.na(latitude), !is.na(longitude),
    # remove duplicates
    !duplicated(latitude)) %>%
  # there are some lat/longs with spaces. Need to fix this
  mutate(longitude = str_remove(longitude, "\\s"), 
         latitude = str_remove(latitude, "\\s")) %>% 
  # convert to numeric
  mutate(longitude = as.numeric(longitude),
         latitude = as.numeric(latitude)) %>% 
  # convert to sf for spatial operations
  st_as_sf(coords = c("longitude", "latitude"),
           crs = 4326,
           remove = FALSE)

# read and clean pictus gbif localities
locs_pic_gbif <- read_tsv(here("analysis", "data", "pictus_gbif.tsv")) %>% 
  dplyr::select(
    genus, 
    species, 
    state = stateProvince,
    longitude = decimalLongitude,
    latitude = decimalLatitude,
    id_code = gbifID
    ) %>% 
  mutate(
    species = str_remove_all(species, "Enyalius "),
    id_code = paste0("pic_", id_code)
    ) %>% 
  st_as_sf(
    coords = c("longitude", "latitude"),
    crs = 4326,
    remove = FALSE
  )

# combine the two datasets
locs_pic <- bind_rows(
  locs_pic, 
  locs_pic_gbif
)

# plot localities
ggplot() +
  geom_sf(data = st_geometry(af), fill = "darkgreen") +
  geom_sf(data = locs_pic) 
```

Spatial thin. I'm using a 20 km buffer. Results in 16 localities.

```{r}
#| label: spthin-pic
set.seed(888273)

#run spthin algorithm. This returns 100 possible combinations of removed localities
output <-
  spThin::thin(
    locs_pic,
    'latitude',
    'longitude',
    'species',
    thin.par = 20,
    reps = 100,
    locs.thinned.list.return = TRUE,
    write.files = FALSE,
    verbose = FALSE
  )

# I want to maximize the # of localities returned  
maxThin <- which(sapply(output, nrow) == max(sapply(output, nrow)))

# if there are multiple iterations with the max # of localities, pick one
maxThin <-
  output[[ifelse(length(maxThin) > 1, maxThin[1], maxThin)]]

# subset locs to match only thinned locs
locs_pic <- locs_pic[as.numeric(rownames(maxThin)), ]

# get the unused locs as a testing data set
# this probably isn't useful since they will overlap heavily with the training data, but seems like another piece of info to look at
test_pic <- locs_pic[-as.numeric(rownames(maxThin)), ]


ggplot() +
  geom_sf(data = st_geometry(af), fill = "darkgreen") +
  geom_sf(data = locs_pic)
```

Write thinned localities to file

```{r}
#| label: write-spthin-pic
#| eval: false

# Write to file
st_write(locs_pic, here("analysis", "output", "thinned_localities", "pictus_thinned.gpkg"))
```

#### Crop environment

First, I need to read in the environmental data.

```{r}
#| label: env-data-pic

bioclims <- terra::rast(here("analysis", "output", "cropped_predictors", "bioclims.tif"))
forest_cover <- terra::rast(here("analysis", "output", "cropped_predictors", "forest_cover.tif"))
```

Cropping and combining variables for analysis.

```{r}
#| label: crop-env-pic
#| warning: false

mcp_pic <- 
  st_convex_hull(st_union(locs_pic)) %>%
  st_buffer(dist = units::set_units(0.5, degree)) %>% 
  terra::vect()


bioclims_pic <- bioclims %>%
  # the bioclims are in a slightly different CRS
  terra::project(forest_cover) %>% 
  terra::crop(mcp_pic) %>% 
  terra::mask(mcp_pic)

forest_cover_pic <- forest_cover %>% 
  terra::crop(bioclims_pic) %>% 
  terra::mask(bioclims_pic[[1]]) 

predictors_pic <- c(bioclims_pic)


plot(bioclims_pic[[1]])
plot(forest_cover_pic)
plot(locs_pic, add = TRUE)
```

#### Predictor correlation

Sample 10000 background points. 4285 actually sampled.

```{r}
#| label: bg-env-pic
set.seed(9745)

# for variable correlations
bg_envt_pic <- terra::spatSample(predictors_pic, 10000, 
                               warn=TRUE, 
                               na.rm = TRUE, 
                               as.df = TRUE,
                               xy = TRUE)

# for use in ENMevaluate
bg_coords_pic <- bg_envt_pic[,c("x", "y")]
```

Next, I'll extract the values for the background points and perform variance inflation factor stepwise selection with a VIF threshold of 10.

```{r}
#| label: vif-pic
# extract values
bg_corr_pic <- bg_envt_pic %>% select(-x, -y)

usdm::vifstep(bg_corr_pic, th=10)
```

The final variable list: BIO4, BIO7, BIO9, BIO13, BIO18, BIO19

```{r}
predictors_pic <- predictors_pic[[c("bio4", "bio7", "bio9", "bio13", "bio18", "bio19")]]
```

#### Maxent model

I'm using a jackknifing model evaluation approach since I only have 15 observations.

```{r}
set.seed(755555598)
coords_pic <- st_coordinates(locs_pic)
colnames(coords_pic) <- c("x", "y")

folds_pic <- ENMeval::get.jackknife(occ = coords_pic, 
                            bg = bg_coords_pic)
```

Run the model. Predictions are clamped to prevent extrapolation. This part takes a little bit. 29 minutes 57 sec on the wall clock on a 2020 M1 MacBook air with 6 cores and 16 GB of memory. Probably because of the high number of background points and the wide range of regularization multipliers I'm exploring.

```{r, eval=FALSE}
set.seed(3767557)

# the vector of regularization multipliers to test
rms <- seq(0.5, 5, 0.5)

# convert the terra object to a raster stack for use in EMNeval
predictors_pic_stack <- raster::stack(predictors_pic)

# iterate model building over all chosen parameter settings
sdm_pic <-
  ENMeval::ENMevaluate(
    occs = coords_pic,
    envs = predictors_pic_stack,
    bg.coords = bg_coords_pic,
    RMvalues = rms,
    fc = c('L', 'LQ'),
    # clamping to prevent model extrapolation
    doClamp = TRUE,
    taxon.name = "pictus",
    partitions = "user",
    user.grp = folds_pic,
    bg.grp,
    clamp = TRUE,
    algorithm = "maxent.jar",
    parallel = TRUE,
    numCores = 6
  )
```

```{r, eval=FALSE}
# write the model to file
write_rds(sdm_pic, here("analysis", "output", "sdm_models", "sdm_pictus.rds"))
```

```{r, echo=FALSE}
sdm_pic <- read_rds(here("analysis", "output", "sdm_models", "sdm_pictus.rds"))
```

##### Model evaluation

Let's take a look at the model results.

```{r}
eval_table_pic <- sdm_pic@results
eval_mods_pic <- sdm_pic@models

names(eval_mods_pic) <-
  str_replace_all(names(eval_mods_pic), "\\.", "\\_")
```

Select the final model. First I'm looking at plots of model evaluation stats to get an idea of the spread of stats.

```{r}
daic_pic <- ggplot(data = eval_table_pic, aes(x = rm, y = delta.AICc, color = fc)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw()

or_pic <- ggplot(data = eval_table_pic, aes(x = rm, y = or.10p.avg, color = fc)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw()

dauc_pic <- ggplot(data = eval_table_pic, aes(x = rm, y = auc.diff.avg, color = fc)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw()

(daic_pic + or_pic) / (dauc_pic)
```

Now I'm going to take a look at tables of delta AICc, omission rate, and AUC to see how close the models are. The model with the lowest AICc has a regularization multiplier of 1 and L feature class. It also has a reasonable omission rate and AUC. I will go forward with this model.

```{r}
eval_table_pic %>% 
  select(delta.AICc, AICc, or.10p.avg, or.mtp.avg, auc.diff.avg, auc.val.avg, rm, fc) %>%
  arrange(delta.AICc) %>% 
  head(10) %>% 
  knitr::kable()
```

Select model.

```{r}
mod_pic <- eval_mods_pic$rm_2_fc_L
opt_seq_tune_pic <- eval_table_pic$tune.args[na.omit(eval_table_pic$delta.AICc) == 0][1]
```

Variable importance. bio7 is the only important variable

```{r}
#| label: var-importance-pic
plot(mod_pic)
```

```{r, eval=FALSE}
png("variable_contribution_pictus.png")
plot(mod_pic)
dev.off()

file.copy(here("variable_contribution_pictus.png"), here("analysis", "output", "sdm_response_curves_variable_importance", "variable_contribution_pictus.png"))
file.remove(here("variable_contribution_pictus.png"))
```

Plot the response curves. Annual range in air temperature is negatively related with presence.

```{r}
#| label: resp-curves-pic
dismo::response(mod_pic)
```

```{r, eval=FALSE}
png("response_curves_pictus.png")
dismo::response(mod_pic)
dev.off()

file.copy(here("response_curves_pictus.png"), here("analysis", "output", "sdm_response_curves_variable_importance", "response_curves_pictus.png"))
file.remove(here("response_curves_pictus.png"))
```

##### Project

I'm projecting the model to the study area extent.

```{r}
pred_pic <- ENMeval::eval.predictions(sdm_pic)[[opt_seq_tune_pic]]
plot(pred_pic)
plot(st_geometry(af), add = TRUE)
plot(st_geometry(locs_pic), pch = 21, bg = alpha("lightgray", 0.5), add = TRUE)
```

```{r}
#| label: proj-cur-pic-thresh
pred_pic_thresh <- pred_pic$rm.2_fc.L
# observed suitabilities
obs_suit_pic <- terra::extract(pred_pic$rm.2_fc.L, locs_pic)
# minimum training presence
min_suit_pic <- min(obs_suit_pic)

pred_pic_thresh[pred_pic_thresh >= min_suit_pic] <- 1
pred_pic_thresh[pred_pic_thresh < min_suit_pic] <- 0

plot(pred_pic_thresh)
plot(st_geometry(af), add = TRUE)
plot(st_geometry(locs_pic), pch = 21, bg = alpha("lightgray", 0.5), add = TRUE)
```

### E. perditus

Note: putative "perditus 2" species localities from Mariana's phylogenetics work have been removed

Reading in data for plotting.

```{r}
# atlantic forest shapefile
af <- read_sf(here("analysis", "data", "atlantic_forest", "atlantic_forest.geojson"))
```

#### Spatial thin

Read and filter localities.

```{r}
locs_per <- read_csv(here("analysis", "data", "enyalius_locs.csv")) %>% 
  # the variable names are messy
  janitor::clean_names() %>%
  filter(species == "perditus",
    !is.na(latitude), !is.na(longitude),
    # remove duplicates
    !duplicated(latitude)) %>%
  # there are some lat/longs with spaces. Need to fix this
  mutate(longitude = str_remove(longitude, "\\s"), 
         latitude = str_remove(latitude, "\\s")) %>% 
  # convert to numeric
  mutate(longitude = as.numeric(longitude),
         latitude = as.numeric(latitude)) %>% 
  # convert to sf for spatial operations
  st_as_sf(coords = c("longitude", "latitude"),
           crs = 4326,
           remove = FALSE)

# plot localities
plot(st_geometry(af))
plot(st_geometry(locs_per), add = TRUE)
```

Spatial thin. I'm using a 20 km buffer. Leaves us with 21 localities.

```{r}
set.seed(117553)

#run spthin algorithm. This returns 100 possible combinations of removed localities
output <-
  spThin::thin(
    locs_per,
    'latitude',
    'longitude',
    'species',
    thin.par = 20,
    reps = 100,
    locs.thinned.list.return = TRUE,
    write.files = FALSE,
    verbose = FALSE
  )

# I want to maximize the # of localities returned  
maxThin <- which(sapply(output, nrow) == max(sapply(output, nrow)))

# if there are multiple iterations with the max # of localities, pick one
maxThin <-
  output[[ifelse(length(maxThin) > 1, maxThin[1], maxThin)]]

# subset locs to match only thinned locs
locs_per <- locs_per[as.numeric(rownames(maxThin)), ]

# get the unused locs as a testing data set
# this probably isn't useful since they will overlap heavily with the training data, but seems like another piece of info to look at
test_per <- locs_per[-as.numeric(rownames(maxThin)), ]

plot(st_geometry(af))
plot(st_geometry(locs_per), add=TRUE)
```

Write thinned localities to file

```{r, eval=FALSE}
# Write to file
st_write(locs_per, here("analysis", "output", "thinned_localities", "perditus_thinned.gpkg"),
         delete_dsn = TRUE)
```

#### Crop environment

First, I need to read in the environmental data.

```{r}
bioclims <- terra::rast(here("analysis", "output", "cropped_predictors", "bioclims.tif"))
forest_cover <- terra::rast(here("analysis", "output", "cropped_predictors", "forest_cover.tif"))
```

Cropping and combining variables for analysis.

```{r, warning=FALSE, message=FALSE}
mcp_per <- st_convex_hull(st_union(locs_per)) %>%
  st_buffer(dist = units::set_units(0.5, degree)) %>% 
  terra::vect()

bioclims_per <- bioclims %>%
  # the bioclims are in a slightly different CRS
  terra::project(forest_cover) %>% 
  terra::crop(mcp_per) %>% 
  terra::mask(mcp_per)

forest_cover_per <- forest_cover %>% 
  terra::crop(mcp_per) %>% 
  terra::mask(mcp_per) 

plot(bioclims_per[[1]])
plot(st_geometry(locs_per), add = TRUE)
plot(forest_cover_per)
plot(st_geometry(locs_per), add = TRUE)

predictors_per <- c(bioclims_per)
```

#### Predictor correlation

Sample 10000 background points. Actually got 4225 points.

```{r}
set.seed(86999784)

# for variable correlations
bg_envt_per <- terra::spatSample(predictors_per, 10000, 
                               warn=TRUE, 
                               na.rm = TRUE, 
                               as.df = TRUE,
                               xy = TRUE)

# for use in ENMevaluate
bg_coords_per <- bg_envt_per[,c("x", "y")]
```

Next, I'll extract the values for the background points and perform variance inflation factor stepwise selection with a VIF threshold of 10.

```{r}
#| label: vif-per
# extract values
bg_corr_per <- bg_envt_per %>% select(-x, -y)

usdm::vifstep(bg_corr_per, th=10)
```

The final variable list: BIO3, BIO7, BIO8, BIO13, BIO18, BIO19

```{r}
predictors_per <- predictors_per[[c("bio3", "bio7", "bio8", "bio13", "bio18",  "bio19")]]
```

#### Maxent model

I'm using a jackknife model evaluation approach since I only have 17 observations.

```{r}
set.seed(42344377)
coords_per <- st_coordinates(locs_per)
colnames(coords_per) <- c("x", "y")

folds_per <- ENMeval::get.jackknife(occ = coords_per, 
                            bg = bg_coords_per)
```

Run the model. Predictions are clamped to prevent extrapolation.

```{r, eval=FALSE}
set.seed(9769997)

# the vector of regularization multipliers to test
rms <- seq(0.5, 5, 0.5)

# convert the terra object to a raster stack for use in EMNeval
predictors_per_stack <- raster::stack(predictors_per)

# iterate model building over all chosen parameter settings
sdm_per <-
  ENMeval::ENMevaluate(
    occs = coords_per,
    envs = predictors_per_stack,
    bg.coords = bg_coords_per,
    RMvalues = rms,
    fc = c('L', 'LQ'),
    # clamping to prevent model extrapolation
    doClamp = TRUE,
    taxon.name = "perditus",
    partitions = "user",
    # going to do a separate run with the final model on testing data
    #occs.testing = test_per,
    user.grp = folds_per,
    clamp = TRUE,
    algorithm = "maxent.jar",
    parallel = TRUE,
    numCores = 6
  )
```

```{r, eval=FALSE}
# write the model to file
write_rds(sdm_per, here("analysis", "output", "sdm_models", "sdm_perditus.rds"))
```

```{r, echo=FALSE}
sdm_per <- read_rds(here("analysis", "output", "sdm_models", "sdm_perditus.rds"))
```

##### Model evaluation

Let's take a look at the model results.

```{r}
eval_table_per <- sdm_per@results
eval_mods_per <- sdm_per@models

names(eval_mods_per) <-
  str_replace_all(names(eval_mods_per), "\\.", "\\_")
```

Select the final model. First I'm looking at plots of model evaluation stats to get an idea of the spread of stats.

```{r}
daic_per <- ggplot(data = eval_table_per, aes(x = rm, y = delta.AICc, color = fc)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw()

or_per <- ggplot(data = eval_table_per, aes(x = rm, y = or.10p.avg, color = fc)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw()

dauc_per <- ggplot(data = eval_table_per, aes(x = rm, y = auc.diff.avg, color = fc)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw()


(daic_per + or_per) / (dauc_per)
```

Now I'm going to take a look at tables of delta AICc, omission rate, and AUC to see how close the models are. The model with the lowest AICc has a relatively high omission error rate (0.24), so I chose the model with a lower OR (0.12) that is only 1.15 AICc away (fc LQ and rm 3.5). It is also a simpler model with a 3.5 regularization multiplier, which is nice.

```{r}
eval_table_per %>% 
  select(delta.AICc, AICc, or.10p.avg, or.mtp.avg, auc.diff.avg, auc.val.avg, rm, fc) %>%
  arrange(delta.AICc) %>% 
  head(10) %>% 
  knitr::kable()
```

Select model.

```{r}
mod_per <- eval_mods_per$rm_2_fc_LQ
opt_seq_tune_per <- eval_table_per$tune.args[eval_table_per$tune.args == "rm.2_fc.LQ"]
```

Variable importance. BIO8 and BIO3.

```{r}
#| label: var-imp-perditus
plot(mod_per)
```

```{r, eval=FALSE}
png("variable_contribution_perditus.png")
plot(mod_per)
dev.off()

file.copy(here("variable_contribution_perditus.png"), here("analysis", "output", "sdm_response_curves_variable_importance", "variable_contribution_perditus.png"))
file.remove(here("variable_contribution_perditus.png"))
```

Plot the response curves. Almost binary response with isothermality, higher = lower suitability. Lower mean daily temps of the wettest quarter correspond with higher suitability.

```{r}
#| label: resp-curves-per
dismo::response(mod_per)
```

```{r, eval=FALSE}
png("response_curves_perditus.png")
dismo::response(mod_per)
dev.off()

file.copy(here("response_curves_perditus.png"), here("analysis", "output", "sdm_response_curves_variable_importance", "response_curves_perditus.png"))
file.remove(here("response_curves_perditus.png"))
```

##### Project

I'm projecting the model to the study area extent.

```{r}
pred_per <- ENMeval::eval.predictions(sdm_per)[[opt_seq_tune_per]]
plot(pred_per)
plot(st_geometry(af), add = TRUE)
plot(st_geometry(locs_per), pch = 21, bg = alpha("lightgray", 0.5), add = TRUE)
```

```{r}
#| label: proj-cur-per-thresh
pred_per_thresh <- pred_per$rm.2_fc.LQ
# observed suitabilities
obs_suit_per <- terra::extract(pred_per$rm.2_fc.LQ, locs_per)
# minimum training presence
min_suit_per <- min(obs_suit_per)

pred_per_thresh[pred_per_thresh >= min_suit_per] <- 1
pred_per_thresh[pred_per_thresh < min_suit_per] <- 0

plot(pred_per_thresh)
plot(st_geometry(af), add = TRUE)
plot(st_geometry(locs_per), pch = 21, bg = alpha("lightgray", 0.5), add = TRUE)
```

### E. iheringii

Reading in data for plotting.

```{r}
#| label: af-shapefile-ihe
# atlantic forest shapefile
af <- read_sf(here("analysis", "data", "atlantic_forest", "atlantic_forest.geojson"))
```

#### Spatial thin

Read and filter localities.

```{r}
#| label: read-locs-ihe
locs_ihe <- read_csv(here("analysis", "data", "enyalius_locs.csv")) %>% 
  # the variable names are messy
  janitor::clean_names() %>%
  filter(species == "iheringii",
    !is.na(latitude), !is.na(longitude),
    # remove duplicates
    !duplicated(latitude)
    ) %>%
  # there are some lat/longs with spaces. Need to fix this
  mutate(longitude = str_remove(longitude, "\\s"), 
         latitude = str_remove(latitude, "\\s")) %>% 
  # convert to numeric
  mutate(longitude = as.numeric(longitude),
         latitude = as.numeric(latitude)) %>% 
  # some localities fall into the ocean when the rasters are aggregated. They need to be bumped by 0.05 degrees to overlap their respective environmental layer. Given the coarseness of the analysis, this does not bias results. 
  mutate(
    longitude = case_when(
      longitude == -47.01956 ~ longitude - 0.05,
      longitude == -47.93278 ~ longitude - 0.05,
      longitude == -47.93139 ~ longitude - 0.05,
      longitude == -48.03889 ~ longitude - 0.05,
      .default = longitude
    ),
    latitude = case_when(latitude == -23.595 ~ latitude + 0.05,
                         .default = latitude)
  ) %>% 
  # convert to sf for spatial operations
  st_as_sf(coords = c("longitude", "latitude"),
           crs = 4326,
           remove = FALSE)

# plot localities
plot(st_geometry(af), ylim = c(min(locs_ihe$latitude), max(locs_ihe$latitude)))
plot(st_geometry(locs_ihe), 
     add = TRUE)
```

Spatial thin. I'm using a 20 km buffer

```{r}
set.seed(2333)

#run spthin algorithm. This returns 100 possible combinations of removed localities
output <-
  spThin::thin(
    locs_ihe,
    'latitude',
    'longitude',
    'species',
    thin.par = 20,
    reps = 100,
    locs.thinned.list.return = TRUE,
    write.files = FALSE,
    verbose = FALSE
  )

# I want to maximize the # of localities returned  
maxThin <- which(sapply(output, nrow) == max(sapply(output, nrow)))

# if there are multiple iterations with the max # of localities, pick one
maxThin <-
  output[[ifelse(length(maxThin) > 1, maxThin[1], maxThin)]]

# subset locs to match only thinned locs
locs_ihe <- locs_ihe[as.numeric(rownames(maxThin)), ]

# get the unused locs as a testing data set
# this probably isn't useful since they will overlap heavily with the training data, but seems like another piece of info to look at
test_ihe <- locs_ihe[-as.numeric(rownames(maxThin)), ]

plot(st_geometry(af))
plot(st_geometry(locs_ihe), add=TRUE)
```

Write thinned localities to file

```{r, eval=FALSE}
# Write to file
st_write(locs_ihe, here("analysis", "output", "thinned_localities", "iheringii_thinned.gpkg"),
         delete_dsn = TRUE)
```

#### Crop environment

First, I need to read in the environmental data.

```{r}
bioclims <- terra::rast(here("analysis", "output", "cropped_predictors", "bioclims.tif"))
forest_cover <- terra::rast(here("analysis", "output", "cropped_predictors", "forest_cover.tif"))
```

Cropping and combining variables for analysis. Cropping to 0.5 degree buffer to accomodate reasonable dispersal

```{r, warning=FALSE, message=FALSE}
mcp_ihe <- st_convex_hull(st_union(locs_ihe)) %>%
  st_buffer(dist = units::set_units(0.5, degree)) %>% 
  terra::vect()

bioclims_ihe <- bioclims %>%
  # the bioclims are in a slightly different CRS
  terra::project(forest_cover) %>% 
  terra::crop(mcp_ihe) %>% 
  terra::mask(mcp_ihe)

forest_cover_ihe <- forest_cover %>% 
  terra::crop(bioclims_ihe) %>% 
  terra::mask(bioclims_ihe) 

plot(bioclims_ihe[[1]])
plot(st_geometry(locs_ihe), add = TRUE)
plot(forest_cover_ihe[[1]])
plot(st_geometry(locs_ihe), add = TRUE)

predictors_ihe <- c(bioclims_ihe)
```

#### Predictor correlation

Sample 10000 background points. Only 6515 were able to be sampled.

```{r}
set.seed(7488)

# for variable correlations
bg_envt_ihe <- terra::spatSample(predictors_ihe, 10000, 
                               warn=TRUE, 
                               na.rm = TRUE, 
                               as.df = TRUE,
                               xy = TRUE)

# for use in ENMevaluate
bg_coords_ihe <- bg_envt_ihe[,c("x", "y")]
```

Next, I'll extract the values for the background points and perform variance inflation factor stepwise selection with a VIF threshold of 10.

```{r}
#| label: vif-ihe
# extract values
bg_corr_ihe <- bg_envt_ihe %>% select(-x, -y)

usdm::vifstep(bg_corr_ihe, th=10)
```

The final variable list: BIO3, BIO7, BIO8, BIO9, BIO14, BIO18

```{r}
predictors_ihe <- predictors_ihe[[c("bio3", "bio7", "bio8", "bio9", "bio14", "bio18")]]
```

#### Maxent model

I'm using a jackknife model evaluation approach since I only have 26 observations.

```{r}
set.seed(3398737)
coords_ihe <- st_coordinates(locs_ihe)
colnames(coords_ihe) <- c("x", "y")

folds_ihe <- ENMeval::get.jackknife(occ = coords_ihe, 
                            bg = bg_coords_ihe)
```

Run the model. Predictions are clamped to prevent extrapolation.

```{r, eval=FALSE}
set.seed(19999923)

# the vector of regularization multipliers to test
rms <- seq(0.5, 5, 0.5)

# convert the terra object to a raster stack for use in EMNeval
predictors_ihe_stack <- raster::stack(predictors_ihe)

# iterate model building over all chosen parameter settings
sdm_ihe <-
  ENMeval::ENMevaluate(
    occs = coords_ihe,
    envs = predictors_ihe_stack,
    bg.coords = bg_coords_ihe,
    RMvalues = rms,
    fc = c('L', 'LQ'),
    # clamping to prevent model extrapolation
    doClamp = TRUE,
    taxon.name = "iheringii",
    partitions = "user",
    # going to do a separate run with the final model on testing data
    #occs.testing = test_ihe,
    user.grp = folds_ihe,
    clamp = TRUE,
    algorithm = "maxent.jar",
    parallel = TRUE,
    numCores = 6
  )
```

```{r, eval=FALSE}
# write the model to file
write_rds(sdm_ihe, here("analysis", "output", "sdm_models", "sdm_iheringii.rds"))
```

```{r, echo=FALSE}
sdm_ihe <- read_rds(here("analysis", "output", "sdm_models", "sdm_iheringii.rds"))
```

##### Model evaluation

Let's take a look at the model results.

```{r}
eval_table_ihe <- sdm_ihe@results
eval_mods_ihe <- sdm_ihe@models

names(eval_mods_ihe) <-
  str_replace_all(names(eval_mods_ihe), "\\.", "\\_")
```

Select the final model. First I'm looking at plots of model evaluation stats to get an idea of the spread of stats.

```{r}
daic_ihe <- ggplot(data = eval_table_ihe, aes(x = rm, y = delta.AICc, color = fc)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw()

or_ihe <- ggplot(data = eval_table_ihe, aes(x = rm, y = or.10p.avg, color = fc)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw()

dauc_ihe <- ggplot(data = eval_table_ihe, aes(x = rm, y = auc.diff.avg, color = fc)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw()

(daic_ihe + or_ihe) / (dauc_ihe)
```

Now I'm going to take a look at tables of delta AICc, omission rate, and AUC to see how close the models are. All of the top models have the same omission rate, similar AUC, and are either LQ or L. I chose the simplest model within 2 AICc of the top model, which was the rm 2.5 and linear feature class model. The top was a rm 2.5 linear quadratic, but even marginally simpler is better within reasonable limits. After skimming through them, they are all very similar models.

```{r}
eval_table_ihe %>% 
  select(delta.AICc, AICc, or.10p.avg, or.mtp.avg, auc.diff.avg, auc.val.avg, rm, fc) %>%
  arrange(delta.AICc) %>% 
  head(10) %>% 
  knitr::kable()
```

Select model.

```{r}
mod_ihe <- eval_mods_ihe$rm_2_fc_LQ
opt_seq_tune_ihe <- eval_table_ihe$tune.args[eval_table_ihe$tune.args == "rm.2_fc.LQ"]
```

Variable importance.

```{r}
#| label: var-importance-ihe
plot(mod_ihe)
```

```{r, eval=FALSE}
png("variable_contribution_iheringii.png")
plot(mod_ihe)
dev.off()

file.copy(here("variable_contribution_iheringii.png"), here("analysis", "output", "sdm_response_curves_variable_importance", "variable_contribution_iheringii.png"))
file.remove(here("variable_contribution_iheringii.png"))
```

Plot the response curves. In order: bio3, bio7^2, bio8, bio9, bio14^2, bio18.

```{r}
#| label: resp-curves-ihe
dismo::response(mod_ihe)
```

```{r, eval=FALSE}
png("response_curves_iheringii.png")
dismo::response(mod_ihe)
dev.off()

file.copy(here("response_curves_iheringii.png"), here("analysis", "output", "sdm_response_curves_variable_importance", "response_curves_iheringii.png"))
file.remove(here("response_curves_iheringii.png"))
```

##### Project

I'm projecting the model to the study area extent. It's not predicting strong suitability for the southern localities. I'll have to look at the genetic structure and locality info more closely to see what's up with them.

```{r}
pred_ihe <- ENMeval::eval.predictions(sdm_ihe)[[opt_seq_tune_ihe]]
plot(pred_ihe)
plot(st_geometry(af), add = TRUE)
plot(st_geometry(locs_ihe), pch = 21, bg = alpha("lightgray", 0.5), add = TRUE)
```

```{r}
#| label: proj-cur-ihe-thresh
pred_ihe_thresh <- pred_ihe$rm.2_fc.LQ
# observed suitabilities
obs_suit_ihe <- terra::extract(pred_ihe$rm.2_fc.LQ, locs_ihe)
# minimum training presence
min_suit_ihe <- min(obs_suit_ihe)

pred_ihe_thresh[pred_ihe_thresh >= min_suit_ihe] <- 1
pred_ihe_thresh[pred_ihe_thresh < min_suit_ihe] <- 0

plot(pred_ihe_thresh)
plot(st_geometry(af), add = TRUE)
plot(st_geometry(locs_ihe), pch = 21, bg = alpha("lightgray", 0.5), add = TRUE)
```
